Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-19 text/paper1/paper.tex
no time for talking

Figure~\ref{fig:gaia_errors} show the Gaia end-of-mission distance and
tangential velocity error estimates for RR Lyrae. The velocity uncertainty depends on the magnitude of the source, which depends on the metallicity: each line in Figure~\ref{fig:gaia_errors} is computed by Monte Carlo sampling from the empirical metallicity distribution of the Galactic halo from \cite{ivezic2008} (centered around [Fe/H] $\sim$ -1.5). Within 2~kpc, Gaia
will be able to measure distances to these stars with better than 2\%
accuracy: RR Lyraes in this volume will be used to test and calibrate
the Spitzer PL relation described above. Beyond the 2~kpc threshold,
the mid-IR PL relation for RR Lyraes will provide more accurate
distance measurements.

\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth, trim=2 2 5 5, clip]{/Users/adrian/projects/streams/plots/paper1/gaia.pdf}
\caption{ todo }\label{fig:gaia_errors}
\end{center}
\end{figure}

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<10$~kpc to 80~kpc. This enormous increase in volume will greatly refine
data on existing debris systems such as the Sagittarius stream.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} illustrates results from the LM10
N-body simulation of dwarf satellite disruption along the expected Sgr
orbit. The particle positions are shown in a coordinate system defined
by the orbit of the satellite orbit \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2).
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ flattening ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right)
\end{align}

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history 
(its mass and stellar populations, orbit, rate of destruction, and
even original distribution in populations [other refs??]). The phase-space distribution of the debris offers a unique probe of
the depth, shape, extent, and orientation of the Milky Way's dark
matter halo that cannot be matched using other techniques on our own
or other galaxies \citep[e.g.,][]{ibata2001}. The LM10 constraints on
Sgr's mass and orbit have invigorated discussions of Sgr in a more
cosmological context, the mass and extent of the original dark matter
halo that hosted it, as well as its effect on the Milky Way itself
\citep{bailin2003,purcell2011,micheldansac2011,gomez2012}.

The combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris opens up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\sim$2~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim$10~km/s;][]{majewski2004}, and ground-based radial
velocities with $\sim$10~km/s accuracy, the measured phase-space
position of each individual RR Lyrae relative to the stream becomes a
powerful probe of the potential (see \ref{fig:}): the next section outlines a new
method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system.

>>>>

Figure~\ref{fig:gaia_errors} show the Gaia end-of-mission distance and
tangential velocity error estimates for RR Lyrae. The velocity uncertainty depends on the magnitude of the source, which depends on the metallicity: each line in Figure~\ref{fig:gaia_errors} is computed by Monte Carlo sampling from the empirical metallicity distribution of the Galactic halo from \cite{ivezic2008} (centered around [Fe/H] $\sim$ -1.5). Within 2~kpc, Gaia
will be able to measure distances to these stars with better than 2\%
accuracy---RR Lyraes in this volume will be used to test and calibrate
the Spitzer PL relation described above. Beyond the 2~kpc threshold,
the mid-IR PL relation for RR Lyraes will provide more accurate
distance measurements.

\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{/Users/adrian/projects/streams/plots/paper1/gaia.pdf}
\caption{ todo }\label{fig:gaia_errors}
\end{center}
\end{figure}

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<$10~kpc to 80~kpc. This enormous increase in volume will greatly refine
data on existing debris systems such as the Sagittarius stream.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} illustrates results from the LM10
N-body simulation of dwarf satellite disruption along the expected Sgr
orbit. The particle positions are shown in a coordinate system defined
by the orbit of the satellite orbit \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history:  
its mass and stellar populations, orbit, rate of destruction, and
even original distribution in populations [other refs??]. The phase-space distribution of the debris offers a unique probe of
the depth, shape, extent, and orientation of the Milky Way's dark
matter halo that cannot be matched using other techniques on our own
or other galaxies \citep[e.g.,][]{ibata2001}. The LM10 constraints on
Sgr's mass and orbit have invigorated discussions of Sgr in a more
cosmological context, the mass and extent of the original dark matter
halo that hosted it, as well as its effect on the Milky Way itself
\citep{bailin2003,purcell2011,micheldansac2011,gomez2012}.

The prospect of combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris opens up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\sim$2~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim$10~km/s;][]{majewski2004}, and ground-based radial
velocities with $\sim$10~km/s accuracy, the measured phase-space
position of each individual RR Lyrae becomes a
powerful probe of the potential (see \ref{fig:}): the next section outlines a new
method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system (Figure~\ref{fig:ps_distance}).

\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{/Users/adrian/projects/streams/plots/paper1/ps_distance.pdf}
\caption{ todo }\label{fig:ps_distance}
\end{center}
\end{figure}
